Legion: Expressing Locality and Independence with Logical Regions

### 2012-08-14 doc/sc2012/circuit.tex
Small edits to the circuit example
Making some colors better in gray scale
Lots of tiny edits for the camera ready version

Listing~\ref{lst:code_ex} shows code for an electrical
circuit simulation, which takes a collection of
wires and nodes where wires meet.  
At each time step the simulation calculates 
currents, distributes charges, and updates voltages.

@@

% FIXME   We add here a description of how the colors print when printed greyscale.
%
Figure~\ref{sfig:part_fig:pvs} shows how the nodes
in a small graph might be split into three pieces.  Blue (lighter) nodes,
attached by wires only to nodes in the same piece, are 
{\em private} to the piece.  Orange (darker) nodes, on the boundary of a
piece, are {\em shared} with (connected to) other pieces.
In the simulation, computations on the
private nodes of different pieces are independent, while
computations on the shared nodes require communication.  To make
this explicit in the program, we partition the nodes region into
private and shared subregions (line 18).  To partition a region, we
provide a {\em coloring}, which is a relation between the elements of
a region and a set of colors.  
%A partition is an object which given a
%coloring and a region $r$ contains for each color $c$ a subregion of
%$r$ containing all the elements of $r$ colored $c$. 
For each color $c$ in the coloring, the partition contains a subregion $r$
of the region being partitioned, with $r$ consisting of the elements colored $c$.  Note that the
partition into shared and private nodes is disjoint because each node has one color.

@@

Lines 31-57 form the bulk of the actual simulation, performing three
passes over the circuit in each iteration.  Each pass loops over an
array of pieces (constructed on lines 27-30 from the partitions),
spawning a task for each piece.  There are no explicit requests for
parallel execution ({\tt spawn} simply indicates a task call and does
not mandate parallel execution) nor is there synchronization 
between the passes.  Which tasks can be run in
parallel within a pass and the required inter-pass dependencies are
determined automatically by the Legion runtime based on the region
access annotations on the task declarations.  The tasks spawned on
lines 32-34 are {\em subtasks} of the main {\tt simulate\_circuit}
task. A subtask can only access regions (or subregions) that its parent task
could access; furthermore, the subtask can only have permissions on a
region compatible with the parent's permissions.  
%This
%requirement plays an important role in our task scheduling algorithm
%(see Section~\ref{sec:soop}).

>>>>

We begin by describing an example program
written in the Legion programming model.
Listing~\ref{lst:code_ex} shows code for an electrical
circuit simulation, which takes a collection of
wires and nodes where wires meet.  
At each time step the simulation calculates 
currents, distributes charges, and updates voltages.

@@

% FIXME   We add here a description of how the colors print when printed greyscale.
%
Figure~\ref{sfig:part_fig:pvs} shows how the nodes
in a small graph might be split into three pieces.  Blue (lighter) nodes,
attached by wires only to nodes in the same piece, are 
{\em private} to the piece.  Red (darker) nodes, on the boundary of a
piece, are {\em shared} with (connected to) other pieces.
In the simulation, computations on the
private nodes of different pieces are independent, while
computations on the shared nodes require communication.  To make
this explicit in the program, we partition the nodes region into
private and shared subregions (line 18).  To partition a region, we
provide a {\em coloring}, which is a relation between the elements of
a region and a set of colors.  
%A partition is an object which given a
%coloring and a region $r$ contains for each color $c$ a subregion of
%$r$ containing all the elements of $r$ colored $c$. 
For each color $c$ in the coloring, the partition contains a subregion $r$
of the region being partitioned, with $r$ consisting of the elements colored $c$.  Note that the
partition into shared and private nodes is disjoint because each node has one color.

@@

Lines 31-57 perform the actual simulation by making three
passes over the circuit for each time step.  Each pass loops over an
array of pieces (constructed on lines 27-30 from the partitions),
spawning a task for each piece.  There are no explicit requests for
parallel execution ({\tt spawn} simply indicates a task call and does
not mandate parallel execution) nor is there explicit synchronization 
between the passes.  Which tasks can be run in
parallel within a pass and the required inter-pass dependencies are
determined automatically by the Legion runtime based on the region
access annotations on the task declarations.  The tasks spawned on
lines 32-34 are {\em subtasks} of the main {\tt simulate\_circuit}
task. A subtask can only access regions (or subregions) that its parent task
could access; furthermore, the subtask can only have permissions on a
region compatible with the parent's permissions.  
%This
%requirement plays an important role in our task scheduling algorithm
%(see Section~\ref{sec:soop}).

### 2012-08-14 doc/sc2012/conclusion.tex
Lots of tiny edits for the camera ready version

We have presented Legion, a programming model for expressing
the abstract locality and independence properties of program
data by using logical regions.  The Legion programming model enables 
a distributed Software-Out-of-Order Processor (SOOP) scheduling 
algorithm.  The SOOP algorithm leverages the locality and independence properties
captured by logical regions to efficiently execute programs.  We also
described a novel mapping interface that guides the SOOP execution
and allows tuning of program performance independent of correctness.
We demonstrated that our SOOP runtime implementation
of Legion was faster than existing implementations of
three benchmarks and that it was able to scale on distributed, heterogeneous
machines.

>>>>

We have presented Legion, a programming model for expressing
the abstract locality and independence properties of program
data by using logical regions.  The Legion programming model enables 
a distributed Software-Out-of-Order Processor (SOOP) scheduling 
algorithm.  The SOOP algorithm leverages the locality and independence properties
captured by logical regions to efficiently execute programs.  We also
described a novel mapping interface that guides the SOOP execution
and allows tuning of program performance independent of correctness.
We demonstrated that for three benchmarks our SOOP runtime 
enabled Legion implementations that were faster than existing codes
and that the SOOP was able to scale on distributed, heterogeneous machines.

### 2012-08-14 doc/sc2012/legion.tex
Lots of tiny edits for the camera ready version

\item We define a {\em software out-of-order processor}, or SOOP, for
  scheduling tasks with region arguments in a manner analogous to how
  out-of-order hardware schedulers process instructions with register
  arguments (Section~\ref{sec:soop}).  In addition to pipelining the execution of tasks over
  several stages, our SOOP is distributed across the machine and is
  also hierarchical and naturally extracts nested
  parallelism (because tasks may recursively spawn subtasks).

@@

\section*{Acknowledgements}
This research used resources of the Keeneland Computing Facility at the
Georgia Institute of Technology, which is supported by the National
Science Foundation under Contract OCI-0910735.  Michael Bauer was supported
by an NVIDIA Graduate Research Fellowship.

>>>>

\item We define a {\em software out-of-order processor}, or SOOP, for
  scheduling tasks with region arguments in a manner analogous to how
  out-of-order hardware schedulers process instructions with register
  arguments (Section~\ref{sec:soop}).  In addition to pipelining the execution of tasks over
  several stages, our SOOP is distributed across the machine and is
  also hierarchical to naturally extract nested
  parallelism (because tasks may recursively spawn subtasks).

@@

\section*{Acknowledgments}
This research used resources of the Keeneland Computing Facility at the
Georgia Institute of Technology, which is supported by the National
Science Foundation under Contract OCI-0910735.  Michael Bauer was supported
by an NVIDIA Graduate Research Fellowship.

### 2012-08-14 doc/sc2012/mapping.tex
Lots of tiny edits for the camera ready version

As mentioned previously, a novel aspect of Legion is the {\em mapping interface},
which gives programmers control over where tasks run and where region instances are placed,
making possible application- or machine-specific mapping decisions that would be difficult for a 
general-purpose programming system to infer.
We describe the interface (Section \ref{sec:mapinterface}),
our base implementation (Section \ref{sec:defmapper}), and the benefits
of creating custom mappers (Section \ref{sec:custommap}).

@@

The mapping interface consists of ten methods that SOOPs call for
mapping decisions.  A {\em mapper} implementing these methods has access to a simple interface
for inspecting properties of the machine, including a list of
processors and their type (i.e. CPU, GPU), a list of
memories visible to each processor, and their latencies and bandwidths.
For brevity we only discuss the three most important interface calls:

@@

\subsection{Custom Mappers}
\label{sec:custommap}
To optimize a Legion program or library, programmers can create one or more custom mappers.  
For each Legion task call the programmer specifies which mapper should be invoked by the runtime for
mapping that particular task.  This allows for the composition of Legion applications and libraries
each with their own custom mappers.

Each custom mapper extends the default mapper.  A programmer need only override the functions 
%necessary for creating a custom mapping.  
he wishes to customize.  Custom mappers can be used to create totally static mappings, 
mappings that memoize their results, or even totally dynamic mappings.  We describe examples of 
custom mappers in Section \ref{sec:exp}.

>>>>

As mentioned previously, a novel aspect of Legion is the {\em mapping interface},
which gives programmers control over where tasks run and where region instances are placed,
making possible application- or machine-specific mapping decisions that would be difficult for a 
general-purpose programming system to infer.  Furthermore, this interface is invoked at runtime which
allows for dynamic mapping decisions based on program input data.
We describe the interface (Section \ref{sec:mapinterface}),
our base implementation (Section \ref{sec:defmapper}), and the benefits
of creating custom mappers (Section \ref{sec:custommap}).

@@

The mapping interface consists of ten methods that SOOPs call for
mapping decisions.  A {\em mapper} implementing these methods has access to a simple interface
for inspecting properties of the machine, including a list of
processors and their type (e.g. CPU, GPU), a list of
memories visible to each processor, and their latencies and bandwidths.
For brevity we only discuss the three most important interface calls:

@@

\subsection{Custom Mappers}
\label{sec:custommap}
To optimize a Legion program or library, programmers can create one or more custom mappers.  
Each custom mapper extends the default mapper.  A programmer need only override the mapper functions 
%necessary for creating a custom mapping.  
he wishes to customize.  Mappers are registered with the runtime and given unique handles.
When a task is launched, the programmer specifies the handle for the mapper that should be invoked by the runtime for
mapping that particular task.

Supporting custom mappers has two benefits.  First, it allows for the composition of Legion applications
and Legion libraries each with their own custom mappers.
Second, custom mappers can be used to create totally static mappings, 
mappings that memoize their results, or even totally dynamic mappings for different 
subsets of tasks in Legion applications.  We describe examples of 
custom mappers in Section \ref{sec:exp}.

%  This allows for the composition of Legion applications and libraries
%each with their own custom mappers.

### 2012-08-14 doc/sc2012/soop.tex
Lots of tiny edits for the camera ready version

Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  A SOOP 
dynamically schedules a stream of tasks in a manner akin to an out-of-order processor scheduling a stream of instructions:
just as an instruction scheduler is constrained by register dependences,
the SOOP is constrained by region dependences.
The SOOP  is pipelined, distributed, and extracts nested parallelism from subtasks.
There are two major challenges in implementing an efficient task scheduler:
\begin{itemize}
\item  For correctness, Legion must preserve data dependences between tasks, which is non-trivial
because the same data may be in multiple different regions or subregions.

>>>>

Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  A SOOP 
dynamically schedules a stream of tasks in a manner akin to an out-of-order processor scheduling a stream of instructions.
Just as an instruction scheduler is constrained by register dependences,
the SOOP is constrained by region dependences.
The SOOP  is pipelined, distributed, and extracts nested parallelism from subtasks.
There are two major challenges in implementing an efficient task scheduler:
\begin{itemize}
\item  For correctness, Legion must preserve data dependences between tasks, which is non-trivial
because the same data may be in multiple different regions or subregions.
